---
title: "HarvardX PH125.9x Data Science: Capstone  \nMovieLens Rating Prediction Project"
author: "Joshua Hendrix"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: false
    includes:
      in_header: pdf_header.tex
  html_document:
    toc: true
    toc_depth: 2
    theme: darkly
---

```{r setup, include=FALSE}

if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")

# Load the cached data if available, otherwise run the full source code.
dataFile <- "prediction_data.RData"
if (file.exists(dataFile)) {
  load(dataFile)
} else {
  source("movielens_Project.R")
  save(
  edx,
  final_holdout_test,
  movieAvg,
  simpleRmse,
  movieEffectRmse,
  userEffectRmse,
  regularizedRmse,
  finalRmse,
  file = "prediction_data.RData")
}
```

# Introduction

Movie or TV show recommendations are an essential feature of most content distribution services, especially streaming services. The recommendation algorithm is an important part of keeping the user engaged by providing personal suggestions and promoting greater satisfaction. The recommendation algorithm also needs to account for several sources of bias when providing a recommendation to the user.

The goal of this project is to develop a recommendation algorithm using statistical analysis and machine learning techniques. The dataset analyzed comes from MovieLens and contains 10 million entries. To evaluate the final prediction model, ten percent of the data will be randomly selected and stored in the `final_holdout_test` variable. The remaining data is split into training and test data sets for model development. Root Mean Squared Error (RMSE) is used as the evaluation metric throughout the algorithm development to measure prediction accuracy.

# Data Analysis

## Data Cleaning
Before moving forward with the analysis, the dataset was examined for missing or invalid values. The `is.na()` function checks for null entries in the dataset.

```{r}
hasNulls <- any(is.na(edx))
hasNulls
```

The data did not contain any missing values therefore no data removal or adjusting was necessary.

## Data Exploration and Visualization

### Summary Statistics

Getting a basic understanding of the data is crucial when developing a prediction model. Basic data analysis and visualizations can help identify potential issues or biases that may influence the final model. A summary statistic printout is a great place to start.

```{r}
summary(edx$rating)
table(edx$rating)
```

```{r fig1, echo=FALSE, fig.align='center', fig.cap="Figure 1: Distribution of Movie Ratings"}
# Plot the distribution of movie ratings.
edx %>%
  ggplot(aes(x = rating)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Movie Ratings", x = "Rating", y = "Number of Ratings")
```

Figure 1 shows a tendency towards higher ratings, with the mean rating around 3.5. Interestingly, the data suggests that users prefer to use whole number ratings over half-step ratings. This could be due to rating systems that only allow integer ratings. Given this data, the mean rating may be a great starting point for the prediction model, helping to estimate how users might rate a movie.

### Impact of Movie Popularity on Rating

Popular movies tend to receive more ratings and often have higher than average ratings. In contrast, movies that are less popular may receive fewer ratings and are likely more biased as a result. By plotting the number of ratings each movie has received versus their average rating, trends can be identified.

```{r fig2, echo=FALSE, message=FALSE, fig.align='center', fig.cap="Figure 2: Impact of Movie Popularity on Rating"}
# Get each movie's average rating
movie_avg_ratings <- edx %>%
  group_by(movieId) %>%
  summarize(avg_rating = mean(rating), count = n())

# Plot movie popularity (number of ratings) vs. average rating
movie_avg_ratings %>%
  ggplot(aes(x = count, y = avg_rating)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick", se = FALSE) +
  scale_x_log10() +
  labs(title = "Number of Ratings vs. Average Rating",
       x = "Number of Ratings", y = "Average Rating")
```

The plot shows a definitive upward average rating trend line as movies receive more ratings. The prediction model will need to take into account the popularity of movies and adjust its rating accordingly.

However, movies with very few ratings (around ten or less) exhibit high variance in their average ratings, which may introduce bias. The prediction model should be cautions to not weigh them too heavily in the prediction process.

### User Bias

User preferences can bias the prediction model by skewing ratings higher or lower than the mean rating. Some users may prefer to rate more negatively where others may prefer to rate more positively. A histogram of user bias can help visualize how much deviation there is from the mean rating.

```{r fig3, echo=FALSE, fig.align='center', fig.cap="Figure 3: User Bias Distribution"}
avg_rating <- mean(edx$rating)
user_bias <- edx %>%
  group_by(userId) %>%
  summarize(user_avg_rating = mean(rating), user_bias = user_avg_rating - avg_rating)

user_bias %>%
  ggplot(aes(x = user_bias)) +
  geom_histogram(binwidth = 0.1, fill = "steelblue", color = "black") +
  labs(title = "User Bias Deviation from Mean Rating", 
       x = "User Bias", y = "Number of Ratings")

```

Reviewing Figure 3, it can be seen that the user ratings deviation from mean rating follow a normal distribution with the mean being close to the movie's mean rating. The mean user bias is `r mean(user_bias$user_bias)` with a standard deviation of `r sd(user_bias$user_bias)`. Using the standard deviation, the data suggests that approximately 68% of users will rate a movie within a half point of the mean, and about 95% of the users will rate within one point of the mean. However, some users deviate from the mean by several standard deviations. Therefore, the prediction model will need to weigh the outliers less than the users that rate closer to the mean rating of movies.







\pagebreak
# Reference

## Figure Code

### Figure 1 Code {#fig1-code}

```{r ref.label="fig1", eval=FALSE}
```

### Figure 2 Code {#fig2-code}

```{r ref.label="fig2", eval=FALSE}
```

### Figure 3 Code {#fig3-code}

```{r ref.label="fig3", eval=FALSE}
```

## External References
The following are references that assisted in creating this rmarkdown file used to generate both the html and pdf documents.

### PDF Figure Position Fix
Figures that could not fit on the remainder of a page would be pushed to the next page. However, the text that followed the figures would be placed on the prior page above the image. This created a difference in the output between the html and pdf formats. A forum post provided the required answer to ensure the figures remained in the same locations as defined in the rmarkdown file.

https://forum.posit.co/t/cant-control-position-of-tables-and-figures-in-knitted-pdf-document/37364